---
title: "Introduction to boostmtree"
format: html
execute:
  freeze: auto
  echo: true
  warning: false
  fig-width: 7
  fig-height: 5
  fig-align: center
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
vignette: >
  %\VignetteIndexEntry{Introduction to boostmtree}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

**boostmtree** extends Friedman's (2001) gradient descent boosting framework to
handle *longitudinal* response data by using multivariate tree base learners and
penalized B-splines (P-splines) to model covariate–time interactions.  The
package supports four response families:

| Family       | Description                                  |
|:-------------|:---------------------------------------------|
| `Continuous` | Real-valued outcome (default)                |
| `Binary`     | 0/1 or two-level factor                      |
| `Nominal`    | Unordered multi-category factor              |
| `Ordinal`    | Ordered multi-category factor                |

Although the package is designed for longitudinal data, it degrades gracefully
to a standard boosting model when each subject has only a single observation.

## Installation

```r
# Released version
install.packages("boostmtree")

# Development version
remotes::install_github("ehrlinger/boostmtree")
```

## Simulating longitudinal data

`simLong()` generates synthetic longitudinal datasets with varying degrees of
covariate–time interaction complexity.  Four simulation models are provided:

| `model` | True formula |
|:-------:|:-------------|
| 0 | `y ~ x1 + x3 + x4` (main effects only) |
| 1 | `y ~ x1 + x3 + x4 + time * x2` (linear interaction) |
| 2 | `y ~ x1 + x3 + x4 + time^2 * x2^2` (quadratic interaction) |
| 3 | `y ~ x1 + x3 + exp(x4) + time^2 * x2^2 * x3` (complex) |

```{r}
#| label: simulate
library(boostmtree)

set.seed(42)
dta <- simLong(
  n      = 50,     # training subjects
  ntest  = 25,     # test subjects
  N      = 5,      # average ~5 time points per subject
  model  = 1,      # linear time × x2 interaction
  family = "Continuous",
  q      = 3       # 3 noise variables added
)

## Data is returned in two formats
## dta$dta   — flat data.frame
## dta$dtaL  — list with $features, $time, $id, $y

cat("Training rows:", sum(dta$dtaL$id %in% unique(dta$dtaL$id)[dta$trn]), "\n")
cat("True model:   ", dta$f.true, "\n")
cat("Covariates:   ", ncol(dta$dtaL$features), "\n")
```

## Fitting a model

The main entry point is `boostmtree()`.  Key tuning parameters are:

| Parameter | Description | Default |
|:----------|:------------|:-------:|
| `M`       | Maximum boosting iterations | 200 |
| `nu`      | Shrinkage / learning rate | 0.05 |
| `K`       | Terminal nodes per tree | 5 |
| `cv.flag` | Use OOB cross-validation for optimal `M` | `FALSE` |
| `nknots`  | Number of B-spline knots | 10 |

```{r}
#| label: fit
## Extract training indices
trn <- dta$trn

fit <- boostmtree(
  x       = dta$dtaL$features[trn, ],
  tm      = dta$dtaL$time[trn],
  id      = dta$dtaL$id[trn],
  y       = dta$dtaL$y[trn],
  family  = "Continuous",
  M       = 50,
  nu      = 0.05,
  cv.flag = TRUE,
  verbose = FALSE
)

print(fit)
```

The `cv.flag = TRUE` option uses the out-of-bag samples at each iteration to
select an optimal stopping point `Mopt`, protecting against overfitting without
a separate held-out validation set.

## Diagnostic plots

`plot.boostmtree()` produces a multi-panel diagnostic display:

```{r}
#| label: plot-fit
plot(fit)
```

Panels include:

- **Fitted vs. observed** values (longitudinal profiles)
- **OOB error rate** vs. boosting iteration with the optimal `M` marked
- **Parameter evolution**: smoothing (`lambda`), variance (`phi`), correlation (`rho`)

## Predicting on new data

```{r}
#| label: predict
pred <- predict(
  fit,
  x  = dta$dtaL$features[-trn, ],
  tm = dta$dtaL$time[-trn],
  id = dta$dtaL$id[-trn],
  y  = dta$dtaL$y[-trn]    # optional — enables test-set RMSE
)

print(pred)
```

When outcomes are provided, `predict()` reports the test-set RMSE (for
Continuous) or misclassification rate (for Binary/Nominal/Ordinal).

```{r}
#| label: plot-pred
plot(pred)
```

## Variable importance

`vimp.boostmtree()` computes permutation-based variable importance.  For grow
objects (training data) it uses OOB samples; for predict objects it uses the
test set.

```{r}
#| label: vimp
vimp_obj <- vimp.boostmtree(fit)
```

`vimpPlot()` displays the results.  For longitudinal models, positive bars
(above the x-axis) represent main effects and negative bars (below) represent
time-interaction effects.

```{r}
#| label: vimp-plot
vimpPlot(vimp_obj)
```

## Marginal and partial dependence plots

### Marginal plots (fast)

`marginalPlot()` bins the covariate into quantile groups and plots the *raw*
(unadjusted) predicted mean at each group.  It is fast and useful for
initial exploration.

```{r}
#| label: marginal
marginalPlot(fit, xvar.names = c("x1", "x2"), plot.it = TRUE)
```

### Partial dependence plots (adjusted)

`partialPlot()` marginalises over all other covariates at each evaluation
point, providing a confounder-adjusted relationship.  It is slower but more
interpretable in the presence of correlated predictors.

```{r}
#| label: partial
partialPlot(fit, xvar.names = c("x1", "x2"), npts = 10)
```

## Binary response

The workflow is identical for binary outcomes.  Simply set
`family = "Binary"` and supply a 0/1 integer or two-level factor as `y`.

```{r}
#| label: binary
set.seed(7)
dta_bin <- simLong(n = 50, ntest = 0, N = 4, model = 1,
                   family = "Binary", q = 2)

fit_bin <- boostmtree(
  x       = dta_bin$dtaL$features,
  tm      = dta_bin$dtaL$time,
  id      = dta_bin$dtaL$id,
  y       = dta_bin$dtaL$y,
  family  = "Binary",
  M       = 50,
  cv.flag = TRUE,
  verbose = FALSE
)

print(fit_bin)
```

## References

Friedman J.H. (2001). Greedy function approximation: a gradient boosting
machine. *Annals of Statistics*, 29(5): 1189–1232.

Friedman J.H. (2002). Stochastic gradient boosting. *Computational Statistics
& Data Analysis*, 38(4): 367–378.

Pande A., Li L., Rajeswaran J., Ehrlinger J., Kogalur U.B., Blackstone E.H.,
Ishwaran H. (2017). Boosted multivariate trees for longitudinal data. *Machine
Learning*, 106(2): 277–305.
doi: [10.1007/s10994-016-5597-1](https://doi.org/10.1007/s10994-016-5597-1)
