---
title: "Longitudinal Data Analysis with boostmtree"
format: html
execute:
  freeze: auto
  echo: true
  warning: false
  fig-width: 7
  fig-height: 5
  fig-align: center
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
vignette: >
  %\VignetteIndexEntry{Longitudinal Data Analysis with boostmtree}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Motivation

Longitudinal studies collect repeated measurements on the same subjects over
time.  The correlation among repeated measures, unbalanced observation
schedules, and the possibility that covariate effects change with time all
present challenges for standard regression methods.

**boostmtree** addresses these challenges by:

1. Using *multivariate tree* base learners that naturally accommodate
   non-linear and interaction effects among covariates.
2. Modeling covariate–time interactions via penalized B-splines, whose
   smoothing parameter is estimated adaptively at each boosting step.
3. Handling unbalanced panels (different numbers of observations per subject)
   without imputation.

This vignette demonstrates the full analysis workflow using the two real
datasets bundled with the package.

---

## Spirometry: Continuous longitudinal response

The `spirometry` dataset contains 9,471 post-lung-transplant FEV1%
measurements from 509 patients along with 23 baseline clinical covariates.

```{r}
#| label: spirometry-load
library(boostmtree)

data(spirometry, package = "boostmtree")

cat("Patients:         ", length(unique(spirometry$id)), "\n")
cat("Total obs:        ", length(spirometry$y), "\n")
cat("Covariates:       ", ncol(spirometry$features), "\n")
cat("FEV1% range:      ", round(range(spirometry$y), 1), "\n")
cat("Time range (yrs): ", round(range(spirometry$time), 2), "\n")
```

### Data structure

`boostmtree()` expects data in *long format* (one row per observation), split
into four components:

| Argument | Description |
|:---------|:------------|
| `x`      | `n_obs × p` covariate matrix (time-invariant per subject) |
| `tm`     | Numeric vector of observation times (length `n_obs`) |
| `id`     | Subject identifier vector (length `n_obs`) |
| `y`      | Response vector (length `n_obs`) |

```{r}
#| label: spirometry-split
set.seed(2024)

## Train/test split by subject
subjects  <- unique(spirometry$id)
n_train   <- floor(0.8 * length(subjects))
trn_subj  <- sample(subjects, n_train)
trn_idx   <- spirometry$id %in% trn_subj

cat("Training subjects:", sum(trn_idx), "\n")
cat("Test subjects:    ", sum(!trn_idx), "\n")
```

The code below shows the full production fit (M = 200, cv.flag = TRUE).  To
keep vignette build time short, pre-computed results are loaded from the
package's `inst/extdata/` directory when available.

```{r}
#| label: spirometry-fit
#| echo: true
#| eval: false
## Full production fit — stored in inst/extdata/fit_spi.rds
fit_spi <- boostmtree(
  x       = spirometry$features[trn_idx, ],
  tm      = spirometry$time[trn_idx],
  id      = spirometry$id[trn_idx],
  y       = spirometry$y[trn_idx],
  family  = "Continuous",
  M       = 200,
  nu      = 0.05,
  K       = 5,
  cv.flag = TRUE,
  verbose = FALSE
)
```

```{r}
#| label: spirometry-fit-load
#| include: false
## Check installed package first, then source tree (local pkgdown builds)
fit_spi_path <- system.file("extdata", "fit_spi.rds", package = "boostmtree")
if (!nzchar(fit_spi_path)) {
  src <- file.path("inst", "extdata", "fit_spi.rds")
  if (file.exists(src)) fit_spi_path <- src
}
if (nzchar(fit_spi_path)) {
  fit_spi <- readRDS(fit_spi_path)
} else {
  ## Fallback: reduced fit when no pre-computed object is available
  fit_spi <- boostmtree(
    x       = spirometry$features[trn_idx, ],
    tm      = spirometry$time[trn_idx],
    id      = spirometry$id[trn_idx],
    y       = spirometry$y[trn_idx],
    family  = "Continuous",
    M       = 50,
    nu      = 0.05,
    K       = 5,
    cv.flag = TRUE,
    verbose = FALSE
  )
}
```

```{r}
#| label: spirometry-print
print(fit_spi)
```

### Diagnostics

```{r}
#| label: spirometry-plot
plot(fit_spi)
```

The top-left panel shows OOB error (RMSE) vs. boosting iteration.  The dashed
vertical line marks the selected `Mopt`.  Use this to judge whether more
iterations (`M`) are needed or if the current budget is sufficient.

### Test-set prediction

```{r}
#| label: spirometry-predict
pred_spi <- predict(
  fit_spi,
  x  = spirometry$features[!trn_idx, ],
  tm = spirometry$time[!trn_idx],
  id = spirometry$id[!trn_idx],
  y  = spirometry$y[!trn_idx]
)

print(pred_spi)
```

### Variable importance

```{r}
#| label: spirometry-vimp
vimp_spi <- vimp.boostmtree(fit_spi)
vimpPlot(vimp_spi)
```

The bar chart separates *main effects* (bars above the axis, labelled in the
upper section) from *time-interaction effects* (bars below the axis).
Variables with both a main effect and a time-interaction effect are the most
interesting: their impact on FEV1% changes over post-transplant time.

### Marginal dependence

```{r}
#| label: spirometry-marginal
## Inspect the top-4 most important variables
top4 <- rownames(vimp_spi[[1]])[order(abs(vimp_spi[[1]][, 1]),
                                       decreasing = TRUE)][1:4]
marginalPlot(fit_spi, xvar.names = top4, plot.it = TRUE)
```

---

## Atrial Fibrillation: Binary longitudinal response

The `AF` dataset tracks weekly AF presence/absence (binary) for 228 patients
enrolled in a surgical ablation trial, with up to 12 months follow-up.

```{r}
#| label: af-load
data(AF, package = "boostmtree")

cat("Patients:    ", length(unique(AF$id)), "\n")
cat("Total obs:   ", length(AF$y), "\n")
cat("Covariates:  ", ncol(AF$feature), "\n")
cat("AF prevalence: ", round(mean(AF$y == 1) * 100, 1), "%\n")
```

### Model fit

For binary responses, set `family = "Binary"`.  The model fits the log-odds on
the linear predictor scale; predictions are returned as probabilities.

```{r}
#| label: af-split
set.seed(2024)

subjects_af <- unique(AF$id)
trn_af_subj <- sample(subjects_af, floor(0.75 * length(subjects_af)))
trn_af      <- AF$id %in% trn_af_subj

cat("Training subjects:", sum(trn_af), "\n")
cat("Test subjects:    ", sum(!trn_af), "\n")
```

The code below shows the full production fit.  Pre-computed results are loaded
from `inst/extdata/fit_af.rds` when available.

```{r}
#| label: af-fit
#| echo: true
#| eval: false
## Full production fit — stored in inst/extdata/fit_af.rds
fit_af <- boostmtree(
  x       = AF$feature[trn_af, ],
  tm      = AF$time[trn_af],
  id      = AF$id[trn_af],
  y       = AF$y[trn_af],
  family  = "Binary",
  M       = 200,
  nu      = 0.05,
  cv.flag = TRUE,
  verbose = FALSE
)
```

```{r}
#| label: af-fit-load
#| include: false
## Check installed package first, then source tree (local pkgdown builds)
fit_af_path <- system.file("extdata", "fit_af.rds", package = "boostmtree")
if (!nzchar(fit_af_path)) {
  src <- file.path("inst", "extdata", "fit_af.rds")
  if (file.exists(src)) fit_af_path <- src
}
if (nzchar(fit_af_path)) {
  fit_af <- readRDS(fit_af_path)
} else {
  ## Fallback: reduced fit when no pre-computed object is available
  fit_af <- boostmtree(
    x       = AF$feature[trn_af, ],
    tm      = AF$time[trn_af],
    id      = AF$id[trn_af],
    y       = AF$y[trn_af],
    family  = "Binary",
    M       = 50,
    nu      = 0.05,
    cv.flag = TRUE,
    verbose = FALSE
  )
}
```

```{r}
#| label: af-print
print(fit_af)
```

```{r}
#| label: af-plot
plot(fit_af)
```

### Test-set prediction

```{r}
#| label: af-predict
pred_af <- predict(
  fit_af,
  x  = AF$feature[!trn_af, ],
  tm = AF$time[!trn_af],
  id = AF$id[!trn_af],
  y  = AF$y[!trn_af]
)

print(pred_af)
```

### Variable importance

```{r}
#| label: af-vimp
vimp_af <- vimp.boostmtree(fit_af)
vimpPlot(vimp_af)
```

### Marginal time trends

The predicted probability of AF over time can be examined by computing
marginal predictions at the observed time points.

```{r}
#| label: af-marginal
marginalPlot(fit_af, xvar.names = colnames(AF$feature)[1:4], plot.it = TRUE)
```

---

## Simulation study: comparing models

`simLong()` supports systematic benchmarking across simulation models.  Below
we compare OOB RMSE across the four simulation scenarios for a continuous
response.

```{r}
#| label: simulation-compare
results <- lapply(0:3, function(m) {
  set.seed(m + 100)
  dta <- simLong(n = 50, ntest = 0, N = 4, model = m,
                 family = "Continuous", q = 2)
  fit <- boostmtree(
    x       = dta$dtaL$features,
    tm      = dta$dtaL$time,
    id      = dta$dtaL$id,
    y       = dta$dtaL$y,
    family  = "Continuous",
    M       = 50,
    cv.flag = TRUE,
    verbose = FALSE
  )
  data.frame(
    model    = paste0("Model ", m),
    formula  = dta$f.true,
    Mopt     = fit$Mopt[1],
    OOB_RMSE = round(min(fit$rmse, na.rm = TRUE), 3)
  )
})

do.call(rbind, results)
```

As expected, error increases with model complexity (richer covariate–time
interactions), but boosting adapts `Mopt` accordingly.

---

## Tuning guide

| Parameter | Effect | Recommendation |
|:----------|:-------|:---------------|
| `M`       | Maximum iterations | Start at 200–500; increase if OOB error is still decreasing at `M` |
| `nu`      | Shrinkage (learning rate) | Smaller values (0.01–0.05) give better results but need larger `M` |
| `K`       | Tree terminal nodes | 3–10; smaller values give smoother fits |
| `nknots`  | B-spline knots for time interaction | 5–15; more knots = more flexible time curves |
| `cv.flag` | OOB-based optimal stopping | Always use `TRUE`; safeguards against overfitting |
| `ntree`   | Trees per base learner (passed to `randomForestSRC`) | 50–200 |

### Learning rate vs. iterations trade-off

A smaller `nu` makes each update more conservative, reducing overfitting risk
but requiring more iterations to converge.  A common strategy is:

```r
# Initial exploration
fit_fast <- boostmtree(x, tm, id, y, M = 100, nu = 0.1, cv.flag = TRUE)

# Refine with slower learning rate
fit_full <- boostmtree(x, tm, id, y,
                       M   = fit_fast$Mopt * 5,
                       nu  = 0.02,
                       cv.flag = TRUE)
```

---

## References

Friedman J.H. (2001). Greedy function approximation: a gradient boosting
machine. *Annals of Statistics*, 29(5): 1189–1232.

Pande A., Li L., Rajeswaran J., Ehrlinger J., Kogalur U.B., Blackstone E.H.,
Ishwaran H. (2017). Boosted multivariate trees for longitudinal data. *Machine
Learning*, 106(2): 277–305.
doi: [10.1007/s10994-016-5597-1](https://doi.org/10.1007/s10994-016-5597-1)

Gillinov A.M. et al. (2015). Surgical ablation of atrial fibrillation during
mitral valve surgery. *New England Journal of Medicine*, 372(15): 1399–1408.

Mason D.P. et al. (2012). Effect of changes in postoperative spirometry on
survival after lung transplantation. *Journal of Thoracic and Cardiovascular
Surgery*, 144: 197–203.
